The Magnetorotational Instability Prefers Three Dimensions

### 2019-11-12 paper/paper.tex
Shortened abstract to comply with PRSA rules.

%%%%%%%%% Insert author address here
\address{$^{1}$Bates College Department of Physics and Astronomy, Lewiston, ME 04240, USA\\
$^{2}$University of Sydney School of Mathematics and Statistics, Sydney, NSW, Australia\\
$^{3}$Statistical Laboratory, DPMMS, University of Cambridge, Cambridge,
UK\\
$^{4}5$Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA\\
$^{5}$Massachusetts Institute of Technology Department of Physics, Cambridge, MA 02139, USA\\
$^{6}$Princeton University Department of Astrophysical Sciences, Princeton, NJ 08544, USA\\
$^{7}$University of Colorado Laboratory for Atmospheric and Space Physics and Department of Astrophysical and Planetary Sciences, Boulder, CO 80309, USA}
\corres{Jeffrey~S.~Oishi\\
\email{joishi@bates.edu}}

\begin{abstract}
The magnetorotational instability (MRI) occurs when a weak magnetic field destabilises a rotating, electrically conducting fluid with inwardly increasing angular velocity.
The MRI is essential to astrophysical disk theory where the shear is typically Keplerian.
Internal shear layers in stars may also be MRI unstable, and they take a wide range of profiles, including near-critical.
We show that the fastest growing modes of an ideal magnetofluid are three-dimensional provided the shear rate, $S$, is near the two-dimensional onset value, $S_c$.
For a Keplerian shear, three-dimensional modes are unstable above $S\approx0.10S_c$, and dominate the two-dimensional modes until $S\approx2.05S_{c}$.
These three-dimensional modes dominate for shear profiles relevant to stars and at magnetic Prandtl numbers relevant to liquid-metal laboratory experiments.
Significant numbers of rapidly growing three-dimensional modes remain well past $2.05S_{c}$. 
These finding are significant in three ways. 
First, weakly nonlinear theory suggests that the MRI saturates by pushing the shear rate to its critical value. 
This can happen for systems, like stars and laboratory experiments, that can rearrange their angular velocity profiles.
Second, the non-normal character and large transient growth of MRI modes should be important whenever three-dimensionality exists.
Finally, three-dimensional growth suggests direct dynamo action driven from the linear instability.
\end{abstract}

>>>>

%%%%%%%%% Insert author address here
\address{$^{1}$Bates College Department of Physics and Astronomy, Lewiston, ME 04240, USA\\
$^{2}$University of Sydney School of Mathematics and Statistics, Sydney, NSW, Australia\\
$^{3}$Statistical Laboratory, DPMMS, University of Cambridge, Cambridge,
UK\\
$^{4}5$Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA\\
$^{5}$Massachusetts Institute of Technology Department of Physics, Cambridge, MA 02139, USA\\
$^{6}$Princeton University Department of Astrophysical Sciences, Princeton, NJ 08544, USA\\
$^{7}$University of Colorado Department of Astrophysical and Planetary Sciences, Boulder, CO 80309, USA}
\corres{Jeffrey~S.~Oishi\\
\email{joishi@bates.edu}}

\begin{abstract}
The magnetorotational instability (MRI) occurs when a weak magnetic field destabilises a rotating, electrically conducting fluid with inwardly increasing angular velocity.
The MRI is essential to astrophysical disk theory where the shear is typically Keplerian.
Shear layers in stars take a wide range of profiles that may be MRI unstable.
We show that the fastest growing modes are 3D when the shear rate, $S$, is near the 2D onset value, $S_c$.
For Keplerian shear, 3D modes are unstable above $S\approx0.10S_c$, and dominate the 2D modes until $S\approx2.05S_{c}$.
Three-dimensional modes dominate for shear profiles relevant to stars and at magnetic Prandtl numbers found in liquid-metal laboratory experiments.
Theory suggests that the MRI saturates by pushing the shear rate to critical in stars and laboratory experiments, where the shear profiles can relax.
whenever three-dimensionality exists.
Finally, 3D growth suggests a laminar dynamo driven by the linear instability.
\end{abstract}
